DATA 621 - Business Analytics and Data Mining

### 2020-11-11 HW4/HW4_Draft.Rmd
Finished this round

# Read and process the test/evaluation file
eval <- fread(paste0(urlRemote, pathGithub, fileTest),
              colClasses = c(rep('integer', 2L), 'double', rep('integer', 4L),
                             'double', 'factor', 'double', rep('factor', 4L),
                             'integer', 'factor', 'double', 'integer',
                             rep('factor', 2L), 'double', 'integer', 'factor',
                             rep('integer', 2L), 'factor'))
eval[, `:=`(INCOME = as.double(gsub('[$,]', '', INCOME)),
          HOME_VAL = as.double(gsub('[$,]', '', HOME_VAL)),
          BLUEBOOK = as.double(gsub('[$,]', '', BLUEBOOK)),
          OLDCLAIM = as.double(gsub('[$,]', '', OLDCLAIM)))]

@@

```{r model1DummySev}
# Model Set 1: Extract those observations with values > 0
trnSev <- trnX[TARGET_AMT > 0, TARGET_AMT]
# Set up dummy variables
trnDumSev <- dummyVars(TARGET_AMT ~ . - INDEX - TARGET_FLAG,
                        fullRank = TRUE, data = trnX[TARGET_AMT > 0])
# Create matrix of dummy values
trnXSD <- predict(trnDumSev, newdata = trnX[TARGET_AMT > 0])
# preProcess with near-zero value and kNN imputation
trnXSDpp <- preProcess(trnXSD, method = c('nzv', 'knnImpute'),
                       k = 3, knnSummary = median)
# Create processed predictors
trnXSDI <- predict(trnXSDpp, newdata = trnXSD)
```
```{r runModel1Sev}
# Model Set 1: Train Model
trC <- trainControl(method = 'none', summaryFunction = defaultSummary,
                    classProbs = FALSE)
set.seed(878)
m1SevFit <- train(x = trnXSDI, y = trnSev, trControl = trC,
                  method = 'glmStepAIC', family = gaussian, direction = 'both',
                  trace = 0)
kable(summary(m1SevFit)$coefficients, digits = 3L,
      caption = "Model 1 Severity Regression Output")
```

@@

```{r validateM1S}
# Model 1 Severity: process validation data similar to training data
tstSev <- tstX[TARGET_AMT > 0, TARGET_AMT]
tstDumSev <- dummyVars(TARGET_AMT ~ . - INDEX - TARGET_FLAG,
                       data = tstX[TARGET_AMT > 0])
tstXSD <- predict(tstDumSev, newdata = tstX[TARGET_AMT > 0])
tstXSDpp <- preProcess(tstXSD, method = c('nzv', 'knnImpute'),
                       k = 3, knnSummary = median)
tstXSDI <- predict(tstXSDpp, newdata = tstXSD)

# Model 1 Severity: predict on validation data
m1Spred <- predict(m1SevFit, newdata = tstXSDI)
# Model 1 Severity: Calculate accuracy statistics
m1Ssum <- postResample(m1Spred, tstSev)
sevCompTable[1, 2] <- m1SRMSE <- m1Ssum[1]
sevCompTable[1, 3] <- m1SMAE <- m1Ssum[3]
sevCompTable[1, 4] <- m1SR2 <- m1Ssum[2]
```

@@

# Accuracy Statistics
m2Ssum <- postResample(m2Spred, tstSev)
sevCompTable[2, 2] <- m2SRMSE <- RMSE(m2Spred, m2.test.r)
sevCompTable[2, 3] <- m2SMAE <- MAE(m2Spred, m2.test.r)
sevCompTable[2, 4] <- m2SR2 <- R2(m2Spred, m2.test.r)
```

@@

# Model 3 Severity: Calculate accuracy statistics
m3Ssum <- postResample(m3Spred, tstSev)
sevCompTable[3, 2] <- m3SRMSE <- m3Ssum[1]
sevCompTable[3, 3] <- m3SMAE <- m3Ssum[3]
sevCompTable[3, 4] <- m3SR2 <- m3Ssum[2]
```

@@

@@

# PREDICTIONS

# CODE APPENDIX
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r loadData}
```
```{r missingVal}
```
```{r zeroVal}
```
```{r statsN}
```
```{r graphsN}
```
```{r graphsF}
```
```{r corrgram}
```
```{r dataClean}
```
```{r trainTestSplit}
```
```{r model1addVars}
```
```{r m2clean}
```
```{r m3clean}
```
```{r model1DummyFreq}
```
```{r model1RunFreq}
```
```{r model1DummySev}
```
```{r model1RunSev}
```
```{r model2RunFreq}
```
```{r model2RunSev}
```
```{r model3RunBaseLog}
```
```{r model3RunFreq}
```
```{r model3RunBaseLin}
```
```{r model3RunSev}
```
```{r setUpCompTables}
```
```{r validateM1F}
```
```{r validateM1S}
```
```{r validateM2F}
```
```{r validateM2S}
```
```{r validateM3F}
```
```{r validateM3S}
```
```{r freqComp}
```
```{r sevComp}
```

>>>>

@@

```{r model1DummySev}
# Model Set 1: Extract those observations with values > 0
trnXS <- trnX[TARGET_AMT > 0,
              ][, `:=`(INDEX = NULL, TARGET_FLAG = NULL)]
trnYS <- trnXS[, TARGET_AMT]

# Set up dummy variables
trnXDum <- dummyVars(TARGET_AMT ~ ., fullRank = TRUE, data = trnXS)

# Create matrix of dummy values
trnXSD <- predict(trnXDum, newdata = trnXS)

# preProcess with near-zero value and kNN imputation using median
trnXDumPP <- preProcess(trnXSD, method = c('nzv', 'knnImpute'), k = 3,
                      knnSummary = median)

# Create processed predictors
trnXSDImp <- predict(trnXDumPP, newdata = trnXSD)
```
```{r runModel1Sev}
# Model Set 1: Train Model
trC <- trainControl(method = 'none', summaryFunction = defaultSummary,
                    classProbs = FALSE)
set.seed(878)
m1SevFit <- train(x = trnXSDImp, y = trnYS, trControl = trC,
                  method = 'glmStepAIC', family = gaussian(),
                  direction = 'both', trace = 0)
kable(summary(m1SevFit)$coefficients, digits = 3L,
      caption = "Model 1 Severity Regression Output")
```

@@

```{r validateM1S}
# Model 1 Severity: process validation data similar to training data
tstXS <- tstX[TARGET_AMT > 0
              ][, `:=`(INDEX = NULL, TARGET_FLAG = NULL)]
tstYS <- tstXS[, TARGET_AMT]
tstXDum <- dummyVars(TARGET_AMT ~ ., data = tstXS, fullRank = TRUE)
tstXSD <- predict(tstXDum, newdata = tstXS)
tstXDumPP <- preProcess(tstXSD, method = c('nzv', 'knnImpute'), k = 3,
                      knnSummary = median)
tstXSDImp <- predict(tstXDumPP, newdata = tstXSD)

# Model 1 Severity: predict on validation data
m1Spred <- predict(m1SevFit, newdata = tstXSDImp)
# Model 1 Severity: Calculate accuracy statistics
m1Ssum <- postResample(m1Spred, tstYS)
sevCompTable[1, 2] <- m1SRMSE <- m1Ssum[1]
sevCompTable[1, 3] <- m1SMAE <- m1Ssum[3]
sevCompTable[1, 4] <- m1SR2 <- m1Ssum[2]
```

@@

# Accuracy Statistics
m2Ssum <- postResample(m2Spred, tstYS)
sevCompTable[2, 2] <- m2SRMSE <- RMSE(m2Spred, m2.test.r)
sevCompTable[2, 3] <- m2SMAE <- MAE(m2Spred, m2.test.r)
sevCompTable[2, 4] <- m2SR2 <- R2(m2Spred, m2.test.r)
```

@@

# Model 3 Severity: Calculate accuracy statistics
m3Ssum <- postResample(m3Spred, tstYS)
sevCompTable[3, 2] <- m3SRMSE <- m3Ssum[1]
sevCompTable[3, 3] <- m3SMAE <- m3Ssum[3]
sevCompTable[3, 4] <- m3SR2 <- m3Ssum[2]
```

@@

**Frequency Model 3**'s performance on the holdout set was the best in two of
the three metrics---accuracy and F1---and will be used to predict the frequency
of a claim on the evaluation set.

@@

**Severity Model 1**'s performance on the holdout set was the best in two of the
three metrics---RMSE and \(R^2\)---and will be used to predict the severity for
claims in the evaluation set for which Model 3 predicts a claim.

However, the performance of this model, or any of the three, is not expected to
be good, as a linear regression with a normal assumption is most often a poor
fit to severity data. Compare the distributions of both training and testing
results for Model 1, ostensibly the best model, on both a real and a log-scale.

```{r badSev, fig.height=4}
m1trnPred <- predict(m1SevFit, trnXSDImp)
densDT <- data.table(trainObs = trnYS,
                     trainPred = m1trnPred,
                     testObs = tstYS,
                     testPred = m1Spred)
densDT <- melt(densDT, variable.factor = FALSE)
densDT[, `:=`(data = ifelse(substr(variable, 1, 5) == 'train', 'train', 'test'),
              len =  lapply(variable, nchar))
              ][, set := ifelse(substr(variable, len, len) == 's',
                                    'observation', 'prediction')
                ][, `:=`(variable = NULL, len = NULL)]
ggplot(densDT, aes(x = value, col = data,  linetype = set)) +
  geom_density(size = 1) + scale_x_continuous(labels = dollar) +
  ggtitle("Density on a Normal Scale")
ggplot(densDT, aes(x = value, col = data, linetype = set)) +
  geom_density(size = 1) + scale_x_log10(labels = dollar) +
  ggtitle("Density on a Log Scale")
```

It is clear that the normal distribution assumption underlying the linear model
is too restrictive. The variance is too low and the tails are too thin. The
log-density plot implies a thicker tailed distribution such as a lognormal or a
Pareto. Nevertheless, we will use Model 3 to predict the losses, understanding
that it is the best option under the artificial constraints of the problem.[^1]

[^1]: Apologies; one of us is an actuary and this is painful 8-) .

# PREDICTIONS
For our group, different model sets had the best performance for frequency and
severity. These models processed the data differently. Therefore, we cannot
simply use the output from Model 3's frequency predictions as the inputs to
Model 1's severity.

What we will do is use Model 3 to determine *which* evaluation observations are
predicted to have claims and then use those observations to predict severity. In
other words, the evaluation data will be processed in accordance with Frequency
Model 3 and the incidents of a claim will be recorded. For those evaluation
observations which return a claim, their raw data will be processed according to
Severity Model 1 to return predicted values.

```{r evalProcessClean}
# Read and process the test/evaluation file
# Can re-use the name DT for convenience now
DT <- fread(paste0(urlRemote, pathGithub, fileTest),
              colClasses = c(rep('integer', 2L), 'double', rep('integer', 4L),
                             'double', 'factor', 'double', rep('factor', 4L),
                             'integer', 'factor', 'double', 'integer',
                             rep('factor', 2L), 'double', 'integer', 'factor',
                             rep('integer', 2L), 'factor'))
neval <- dim(DT)[[1]]
DT[, `:=`(INCOME = as.double(gsub('[$,]', '', INCOME)),
          HOME_VAL = as.double(gsub('[$,]', '', HOME_VAL)),
          BLUEBOOK = as.double(gsub('[$,]', '', BLUEBOOK)),
          OLDCLAIM = as.double(gsub('[$,]', '', OLDCLAIM)))]
setattr(DT$MSTATUS, 'levels', c('Yes', 'No'))
setattr(DT$SEX, 'levels', c('M', 'F'))
setattr(DT$EDUCATION, 'levels', c('<HighSchool', 'Bachelors', 'Masters', 'PhD',
                                  'HighSchool'))
setattr(DT$JOB, 'levels', c('Unknown', 'Clerical', 'Doctor', 'HomeMaker',
                            'Lawyer', 'Manager', 'Professional', 'Student',
                            'BlueCollar'))
setattr(DT$CAR_TYPE, 'levels', c('Minivan', 'PanelTruck', 'Pickup',
                                 'SportsCar', 'Van', 'SUV'))

As there are `r neval` observations, we will not printing them in this document.
Rather, we will show an estimated frequency rate. The actual observations for
which claims are predicted may be obtained from the appropriate data object in
the code, as found in the CODE APPENDIX.

## Evaluation Data Processing and Cleaning
In this section, the evaluation set will be processed and cleaned as the per the
training set. The model-specific imputations, variable creations, and
predictions will be done in their own sections.

## Frequency
Missing values will be imputed, with `INDEX` and the two prediction targets
ignored, of course. Once done, the claim incidents will be predicted.
```{r evalPredictFreq}
# Impute the missing values but not INDEX or the values we will be predicting!
set.seed(94)
m3eval <- complete(mice(data = DT %>%
                          select(-c(INDEX, TARGET_FLAG, TARGET_AMT)),
                        method = "pmm", print = FALSE), 3)
evalFreqFit <- predict(m3FreqFit, m3eval)
numPred <- sum(evalFreqFit == "Claim")
claimRate <- numPred / neval
claimIDX <- which(evalFreqFit == "Claim")
```

The model predicts `r numPred` claims out of `r neval` observations for a claim
rate of `r prettyNum(claimRate * 100, digits = 3L)`%. The actual outcomes are
stored in the `evalFreqFit` object in the code.

## Severity
The individual `r numPred` observations for which Model 3 predicted a claim will
be extracted from the state of the evaluation set prior to Model 3-specific
adjustments, processed per Severity Model 1, and then the claim values will be
predicted.

Once again, with so many observations, we will not simply print all `r numPred`.
Rather, we will provide some summary statistics and graphs of the distribution
of the claims. The actual claim values can be extracted from the appropriate
data structure in the code, of course.

```{r evalPredictSev, fig.height=4}
# Extract the "claim" observations from the data
evalObs <- DT[claimIDX, ]

# At this point the unneeded columns of INDEX and the 2 targets will be removed
evalObs[, `:=`(INDEX = NULL, TARGET_FLAG = NULL, TARGET_AMT = NULL)]

# Create the OWN variable
evalObs[, OWN := factor(ifelse(HOME_VAL > 0, 1, 0), levels = c(0, 1),
                        labels = c('No', 'Yes'))]

# Set up dummy variables
evalDumSev <- dummyVars( ~ ., fullRank = TRUE, data = evalObs)
# Create matrix of dummy values
evalDumSevMatrix <- predict(evalDumSev, newdata = evalObs)
# preProcess with near-zero value and kNN imputation
evalDumSevPP <- preProcess(evalDumSevMatrix, method = c('nzv', 'knnImpute'),
                           k = 3, knnSummary = median)
# Create matrix of processed and imputed predictors
evalDumSevPPI <- predict(evalDumSevPP, newdata = evalDumSevMatrix)
evalSevFit <- predict(m1SevFit, evalDumSevPPI)

# Summary Statistics
tabularComp <- data.table(Set = c("Training", "Testing", "PredictedEval"),
                          Min = c(min(trnYS), min(tstYS), min(evalSevFit)),
                          Median = c(median(trnYS), median(tstYS),
                                     median(evalSevFit)),
                          Max = c(max(trnYS), max(tstYS), max(evalSevFit)),
                          Mean = c(mean(trnYS), mean(tstYS), mean(evalSevFit)),
                          SD = c(sd(trnYS), sd(tstYS), sd(evalSevFit)))
tabularComp[, CV := SD / Mean, by = Set]

ggplot(data.frame(x = evalSevFit), aes(x = x)) + geom_density() +
  scale_x_continuous(labels = dollar)

knitr::kable(tabularComp, digits = 3L,
             caption = "Severity Statistics Comparison")
```

The actual outcomes are stored in the `evalSevFit` object in the code. The table
and the graph show how the predictions are ***much*** more constrained than
either the training or testing sets. The means are relatively close, which is
to be expected when trying to shoehorn non-normal data into a Gaussian
framework.

# CODE APPENDIX
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r loadData}
```
```{r missingVal}
```
```{r zeroVal}
```
```{r statsN}
```
```{r graphsN}
```
```{r graphsF}
```
```{r corrgram}
```
```{r dataClean}
```
```{r trainTestSplit}
```
```{r model1addVars}
```
```{r m2clean}
```
```{r m3clean}
```
```{r model1DummyFreq}
```
```{r model1RunFreq}
```
```{r model1DummySev}
```
```{r model1RunSev}
```
```{r model2RunFreq}
```
```{r model2RunSev}
```
```{r model3RunBaseLog}
```
```{r model3RunFreq}
```
```{r model3RunBaseLin}
```
```{r model3RunSev}
```
```{r setUpCompTables}
```
```{r validateM1F}
```
```{r validateM1S}
```
```{r validateM2F}
```
```{r validateM2S}
```
```{r validateM3F}
```
```{r validateM3S}
```
```{r freqComp}
```
```{r sevComp}
```
```{r badSev}
```
```{r evalProcessClean}
```
```{r evalPredictFreq}
```
```{r evalPredictSev}
```
